\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-20 submission/manuscript.Rmd
update manu

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two (i.e., the subjective reaction associated with the experience of a given stimulus), contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence - the extent to which one feels like "being there", as if the object of experience was real - when engaged in a fictional movie or a VR environment [@sanchez2005presence; @makowski2017being]. Conversely, beliefs that emotional stimuli were fake (e.g., that emotional scenes were not authentic but instead involved actors and movie makeup) were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated  with greater belief [@martel2020reliance; @bago2022emotion]. Additionally, other factors, such as the stimulus' perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in our processing and reaction to real as opposed to non-real material.

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of faces evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, some studies report that when the nature of the faces was ambiguous, artificially created faces that were previously rated as more attractive were judged by subjects to be less real [@tucciarelli2020]. However, as the attractiveness ratings were given by independent raters instead of the participants, the direction of the relationship between perceived realness and attractiveness cannot be concluded. To this end, @liefooghe2022faces reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that reality beliefs have an effect on face attractiveness ratings, the opposite question, whether attractiveness could drive simulation monitoring, has received little attention to date.

<!-- Goals and hypotheses -->
This study primarily aims at exploring the effect of face attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link of dispositional traits, such as personality and attitude towards AI, with inter-individual simulation monitoring tendencies. Note that the discriminative accuracy between "true" photos and "true" artificially-generated images is not relevant for this study, which focuses on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

In the first part of the study, participants answered a series of personality questionnaires, including the *Mini-IPIP6* [24 items, @sibley2011] measuring 6 personality traits, the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development] assessing social anxiety levels, the *FFNI-BF* [30 items, @jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS* [18 items, @freeman2021revised] measuring 2 dimensions related to paranoid thinking; the *IUS-12* [12 items, @carleton2007fearing] measuring intolerance to uncertainty. Finally, we created 5 items pertaining to expectations about AI-generated images technology (**TODO: write here some of the questions**). To lower their saliency and possibly prime the subjects about the task, we mixed these items with 5 items of the attitudes towards AI scale [*GAAIS*, @schepman2020initial]. This scale was presented after the social anxiety questionnaires, and 3 attention check questions were embedded in the questionnaires.

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the validated American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order. Following each stimulus, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales. (**TODO: add details and justifications**)

In the last part of the study, participants were informed that about half of the face images previously seen were AI-generated (the instructions used a cover story mentioning that the research was aimed at validating a new face generation algorithm). The same set of stimuli was displayed again for 500 ms in a new randomized order. This time, after each display, participants were asked to express their belief regarding the nature of the stimulus using a visual analog scales (with *Fake* and *Real* as the two extremes). The study was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

<!-- Attractiveness and Beauty -->

Although attractiveness did not seem to be the primary drive underlying simulation monitoring of face images, we do nonetheless report significant associations, with a different pattern depending on the participant's gender. The quadratic relationship found for female participants is aligned with our hypothesis that salient faces (i.e., rated as very attractive or very unattractive) are judged to be more real. Alternatively, we could interpret the mostly positive linear relationship for males under an evolutionary lens. In particular, males purportedly place more emphasis on facial attractiveness as a sign of reproductive potential, as compared with females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender; @fink2006facial]. As such, it is possible that the evolutionary weight associated with attractiveness skewed the perceived saliency towards attractive faces, rendering attractive faces are significantly more salient than unattractive ones, in turn distorting the relationship with simulation monitoring. Future studies should test this saliency-based hypothesis by measuring constructs closer to salience and its effects, for instance using neuroimaging (**ref on saliency in EEG or fMRI**) or physiological markers [in particular, heart rate deceleration, **include refs**].

<!-- Beauty -->
discuss why the link with beauty is slightly different. Note that for models for beauty we adjusted the models for trustowrthiness and familiarity, hence possibly masking the multidimensionality of attractiveness.

<!-- trustworthiness -->
In contrast to the findings for attractiveness, trustworthiness linearly increased beliefs in realness for females only. Given the evidence for the effect in the opposite direction, suggesting that faces presented as were rated as more trustworthy [@liefooghe2022faces; @hoogers2021effect; @balas2017], we expected such link to be present for both genders. One of the contributing mechanism to this dimorphism could have been the increased risk-taking aversion reported in females [explained evolutionary as a compromise their reproductive potential, @van2020sex], to which perceived facial trustworthiness relates [@hou2019survival]. However, faces judged as highly untrustworthy should have appear as even more salient (representing an evolutionary threat), and hence be judged as more real. As such, further studies are needed to investigate the causes of the increased simulation monitoring sensitivity to trustworthiness in females.

<!-- familiarity -->
Contrary to our hypothesis, familiarity was not found to be significantly related to simulation monitoring decisions. It is to note that the distribution of familiarity ratings was strongly skewed, and that a low number of pictures was rated as highly familiar. Future studies could re-examine this aspect by experimentally manipulating familiarity, for instance by modulating the amount of exposures of items before the simulation monitoring ratings.

<!-- inter-indiv correlates (HH- negative with confidence in real; positive with acclaim-seeking and grandiose; Acc-seeking - positive with confidence in fake) -->

<!-- AI attitudes (enthusiam - positive with confidence in both real and fake judgements) -->

<!-- Re-exposure Delay -->

Although the order of presentation of the facial images was randomized to reduce effects of adaptation, re-exposure delay was found to have a significant negative effect on simulation monitoring decisions, with shorter delays being associated with faces being rated as more real. Interestingly, while it could be posited that shorter delays led to the faces appearing more familiar and thereby increased people's belief in its realness, this seems unlikely considering perceived familiarity did not significantly affect simulation monitoring decisions made, even after controlling for attractiveness and trustworthiness. Alternatively, shorter re-exposure delays could have led to the faces being better remembered, thus triggering autobiographical memory processes [@gobbini2013prioritized] and evoking a sense of personal relevance during the repeated display. Indeed, fictional stimuli that were associated with more personal memories have been shown to up-regulate emotions [@sperduti2016paradox; @makowski2017being], thus biasing the realness of the given stimulus

>>>>

<!-- Reactive features -->
Beyond stimulus- and individual-related characteristics, evidence suggests that the interaction between the two (i.e., the subjective reaction associated with the experience of a given stimulus), contributes to simulation monitoring decisions. For instance, the intensity of experienced emotions have been shown to increase one's sense of presence - the extent to which one feels like "being there", as if the object of experience was real - when engaged in a fictional movie or a VR environment [@sanchez2005presence; @makowski2017being]. Conversely, beliefs that emotional stimuli were fake (e.g., that emotional scenes were not authentic but instead involved actors and movie makeup) were found to result in emotion down-regulation [@makowski2019phenomenal; @sperduti2017distinctive]. In line with these findings, studies on susceptibility to fake news have also found heightened stimulus emotionality to be associated  with greater belief [@martel2020reliance; @bago2022emotion]. Additionally, other factors, such as the stimuli's perceived self-relevance [@sperduti2016paradox; @goldstein2009pleasure], as well as familiarity [@begg1992dissociation], could also play a role in our processing and reaction to real as opposed to non-real material.

<!-- Attractiveness -->
AI-generated images of faces, due to their popularity as a target of CGI technology and the possibility of experimentally manipulating facial features, are increasingly used to study face processing as related to saliency or emotions, as well as to other important components of face evaluation, such as trustworthiness or attractiveness [@sobieraj2014beautiful; @balas2017; @tsikandilakis2019beauty; @calbi2017]. Interestingly, some studies report that when the nature of the faces was ambiguous, artificially created faces that were previously rated as more attractive were judged by subjects to be less real [@tucciarelli2020]. However, as the attractiveness ratings were given by independent raters instead of the participants, the direction of the relationship between perceived realness and attractiveness cannot be concluded. To this end, @liefooghe2022faces reports that attractiveness ratings differed significantly between participants who were told that the faces were AI-generated from those who had no prior knowledge. Whereas this line of evidence suggests that reality beliefs have an effect on face attractiveness ratings, the opposite question, whether attractiveness could drive simulation monitoring, has received little attention to date.

<!-- Goals and hypotheses -->
This study primarily aims at exploring the effect of facial attractiveness on simulation monitoring, i.e., on the beliefs that an image is real or artificially generated. Based on the embodied reality theory [outlined in @makowski2018cognitive; @makowski2019phenomenal], which suggests that salient and emotional stimuli are perceived to be more real, we hypothesize a quadratic relationship between perceived realness and attractiveness: faces rated as highly attractive or unattractive will more likely be believed to be real. We expect a similar relationship with trustworthiness ratings given its well-established link with attractiveness [@liefooghe2022faces; @bartosik2021you; @garrido2017kdef; @little2011facial], and a positive relationship with familiarity (as more familiar faces would appear as more salient, self-relevant and anchored in reality). Additionally, we will further explore the link of dispositional traits, such as personality and attitude towards AI, with inter-individual simulation monitoring tendencies. Note that the discriminative accuracy between "true" photos and "true" artificially-generated images is not relevant for this study, which focuses on the beliefs that a stimulus is real or fake, independently of its true nature.

@@

In the first part of the study, participants answered a series of personality questionnaires, including the *Mini-IPIP6* [24 items, @sibley2011] measuring 6 personality traits, the *SIAS-6* and the *SPS-6* [6 items each, @peters2012development] assessing social anxiety levels, the *FFNI-BF* [30 items, @jauk2022validation] measuring 9 facets of narcissism; the *R-GPTS* [18 items, @freeman2021revised] measuring 2 dimensions related to paranoid thinking; and the *IUS-12* [12 items, @carleton2007fearing] measuring intolerance to uncertainty. Finally, we devised 5 items pertaining to expectations about AI-generated images technology ("I think current Artificial Intelligence algorithms can generate very realistic images"). To lower their saliency and the possibility of it priming the subjects about the task, we mixed these items with 5 items from the general attitudes towards AI scale [*GAAIS*, @schepman2020initial]. This scale was presented after the social anxiety questionnaires, and 3 attention check questions were embedded in the set of scales.

In the second part of this study, 109 photos of neutral-expression faces of real individuals from the validated American Multiracial Face Database (AMFD, [@chen2021broadening]) were presented to the participants for 500ms each, in a randomized order, following a fixation cross display for 750ms. After each stimulus presentation, ratings of **Attractiveness** ("I find this person attractive"), **Beauty** ("This face is good-looking"), **Trustworthiness** ("I find this person trustworthy") and **Familiarity** ("This person reminds me of someone I know") were collected using visual analog scales. Notably, as facial attractiveness is a multidimensional construct, encompassing evolutionary, sociocultural, biological as well as cognitive aspects [@rhodes2006evolutionary; @han2018beauty], we assessed attractiveness using 2 scales, measuring general attractiveness (**Attractiveness**) and physical attractiveness (**Beauty**), to determine whether the effects on simulation monitoring are driven by physical features.

In the last part of the study, participants were informed that about half of the face images previously seen were AI-generated (the instructions used a cover story mentioning that the research was aimed at validating a new face generation algorithm). The same set of stimuli was displayed again for 500 ms in a new randomized order. This time, after each display, participants were asked to express their belief regarding the nature of the stimulus using a visual analog scale (with *Fake* and *Real* as the two extremes). The study was implemented using *jsPsych* [@de2015jspsych], and the full set of instructions is available in the experiment code.

@@

<!-- Attractiveness -->
Although attractiveness did not seem to be the primary drive underlying simulation monitoring of face images, we do nonetheless report significant associations, with a different pattern observed depending on the participant's gender. In particular, the quadratic relationship found for female participants is aligned with our hypothesis that salient faces (i.e., rated as very attractive or very unattractive) are judged to be more real. Alternatively, we could interpret the mostly positive linear relationship for males under an evolutionary lens. Specifically, males purportedly place more emphasis on facial attractiveness as a sign of reproductive potential, as compared with females, who tend to value characteristics signaling resource acquisition capabilities [@qi2022gender; @fink2006facial]. However, while it is possible that the evolutionary weight associated with attractiveness skewed the perceived saliency towards attractive faces, rendering attractive faces as significantly more salient than unattractive ones and in turn distorting the relationship with simulation monitoring, beauty, operationalized as physical attractiveness, was found to have a significant positive linear relationship with simulation monitoring decisions for males even after adjusting for the effects of trustworthiness and familiarity; both characteristics that are evolutionarily advantageous [@hou2019survival; @little2011facial]. As such, future studies are warranted to test this saliency-based hypothesis by measuring constructs closer to salience and its effects, for instance using neuroimaging [@lou2015perceptual; @indovina2007dissociation] or physiological markers [in particular, heart rate deceleration, @goldstein2002influence].

<!-- trustworthiness -->
In contrast to the findings for attractiveness, trustworthiness linearly increased beliefs in realness for females only. Given the evidence suggesting that faces presented as real were rated as more trustworthy [@liefooghe2022faces; @hoogers2021effect; @balas2017], we expected such link to be present for both genders. One of the contributing mechanism to this dimorphism could have been the increased risk-taking aversion reported in females [explained evolutionarily as a compromise to their reproductive potential, @van2020sex], to which perceived facial trustworthiness relates [@hou2019survival]. That being the case, however, faces judged as highly untrustworthy should have appeared as even more salient (representing an evolutionary threat), and hence be judged as more real. As such, further studies are needed to investigate the causes of the increased simulation monitoring sensitivity to trustworthiness in females.

<!-- familiarity -->
Contrary to our hypothesis, familiarity was not found to be significantly related to simulation monitoring decisions. However, it is to note that the distribution of familiarity ratings was strongly skewed, and that a low number of pictures was rated as highly familiar. Future studies could re-examine this aspect by experimentally manipulating familiarity, for instance by modulating the amount of exposures of items before the simulation monitoring ratings.

<!-- inter-indiv correlates -->
Unsurprisingly, participants with higher scores in Narcissism traits of Acclaim Seeking and Grandiose Fantasies were generally more confident in their simulation monitoring decisions. While early researchers had proposed narcissism to be a unitary construct [@raskin1988principal; @sedikides2004normal], recent studies cast narcissism as encompassing 2 unique dimensions - namely narcissistic vulnerability, which is characterized by low self-esteem and a need for admiration, and narcissistic grandiosity, which encompasses high self-esteem, arrogance and aggressive self-assertiveness [@pincus2011narcissistic]. Whereas narcissism has long been linked to over-confidence in decision-making [@campbell2004narcissism], a growing body of research indicates that this relationship is likely mostly driven by facets of narcissistic grandiosity (such as Acclaim Seeking), which drives individuals to see themselves as more superior than others and to externalize blame for poor decision-making consequences [@o2021grandiose]. Indeed, the evidence alludes that attributes specific to narcissistic grandiosity predict individuals' overconfidence in cognitive decision-making [@chatterjee2017master; @brunell2017narcissism; @o2021grandiose].

Conversely, the literature on the correlation between honesty-humility and perceptual confidence judgements remain comparatively scarce, albeit some existing research links higher levels of honesty-humility to increased perceptions of risk and greater risk aversion [@levidi2022understanding; @weller2012honest]. As such, it could be argued that individuals with higher levels of honesty-humility could inherently possess a greater simulation monitoring sensitivity, perceiving a higher risk involved in making decisions on realness and therefore be less confident of their beliefs of reality. This is further supported by convergent evidence finding a negative link between honesty-humility and narcissism [@hodson2018dark]. Therefore, further research examining the relationship between simulation monitoring decisions, personality traits, particularly that related to the dark triad and honesty-humility, and the perceived risks involved in making such evaluations, is warranted.

<!-- AI attitudes -->
Interestingly, despite the ubiquity of AI, there is a lack of research pertaining to the influence of people's AI attitudes on their confidence in their discriminative ability to distinguish between output that is "true" (i.e., images of real faces) from those that is artificially-generated. One possible explanation for the positive relationship observed between AI-Enthusiasm and confidence in reality judgements could be that individuals with highly positive AI attitudes may perceive themselves as having greater knowledge about AI and its capabilities, hence permitting them to be more confident in their simulation monitoring decisions. This is in line with previous research findings, which indicate that greater enthusiasm about AI is related to having a higher confidence in one's knowledge regarding AI competencies [@said2022artificial]. Moreover, @said2022artificial further reports that AI attitudes interacts with people's perceived self-knowledge to influence their perception of the opportunities and risks accorded by AI applications. Specifically, people with AI attitudes that were highly positive and that were more confident in their AI knowledge were more likely to over-estimate the advantages of AI and under-estimate the risks involved in their application. As such, it is plausible that individuals' perceived self-knowledge about AI could have modulated the relationship between their attitudes towards AI and their confidence in their reality judgements.

<!-- Re-exposure Delay -->
Although the order of presentation of the facial images was randomized to reduce effects of adaptation, re-exposure delay was found to have a significant negative effect on simulation monitoring decisions, i.e., shorter delays were associated with faces being rated as more real. This may be because shorter re-exposure delays could have led to the faces being better remembered and appearing more familiar than faces that were displayed after a longer delay, thereby triggering self-referential and autobiographical memory processing during the repeated display [@abraham2009reality; @gobbini2013prioritized; @taylor2009neural]. Although the distribution variance in re-exposure delay is small, recent research using event-related brain potentials report representations of faces to be qualitatively similar regardless of the degree and type of familiarity (e.g., personal or media-based), only distinguishing between familiar and unfamiliar faces [@wiese2021familiarity]. Moreover, fictional stimuli has been shown to up-regulate emotions when associated with familiarity [@sperduti2016paradox; @makowski2017being], thus biasing its salience and perceived realness. Future studies could hence further investigate the modulatory effects of types and degrees of familiarity on perceived realness judgements.

In conclusion, research on our ability to discriminate between what is real from what is not is increasingly relevant given the current expansion on technologies related to virtual and augmented realities. The aim of the present study was to examine whether facial attractiveness can significantly influence our simulation monitoring decisions. In line with our hypothesis, we found faces rated as attractive and unattractive to be perceived as more real for females as a result of an increased salience. However, for males, the observation of a positive linear relationship between attractiveness and beauty with simulation monitoring decisions, adjusted for evolutionarily advantageous features such as trustworthiness, suggest a more complex relationship than hypothesized. As such, future studies could further investigate the role of attractiveness in simulation monitoring, taking greater consideration of the former's multidimensionality.
